\chapter{Introduction}\label{c:introduction}

Recovering the three-dimensional structure of a dynamic scene from visual observations is a
foundational challenge in computer vision with direct implications for robotics and autonomous
navigation~\cite{hartley2004multiple}. Recent spatial foundation models---such as
MASt3R-SfM~\cite{mast3r}, VGGT~\cite{vggt}, PI3, and PI3X~\cite{pi3}---have achieved
remarkable per-frame geometric fidelity by regressing dense 3D pointmaps directly from images.
Yet in 4D settings, where temporal coherence across a video sequence is as critical as
spatial accuracy, these models fail: assembled frame-by-frame, their outputs exhibit severe
high-frequency jitter that renders them unsuitable for robotics applications requiring smooth,
metric-consistent depth. Geometry-refinement approaches such as GGPT~\cite{ggpt} attempt to
correct these inconsistencies by grounding spatial predictions with sparse SfM guidance, but
fail to generalise to sparse-view dynamic scenes, worsening Chamfer distance by
3--10$\times$. Conversely, dedicated 4D architectures---VGGT4D~\cite{vggt4d} and
MonoFusion~\cite{monofusion}---sacrifice geometric precision to buy temporal smoothness,
inflating Chamfer distance by 20--160\,\% and producing systematic alignment failures in
sparse-view settings.

The central finding of this thesis is that a \emph{decoupled} paradigm outperforms both
families: take a strong per-frame 3D estimator, align its outputs into a common coordinate
frame through a lightweight temporal alignment strategy, and apply a confidence-weighted,
mask-aware Gaussian post-processing filter (Opt) to enforce static-scene consistency.
This approach reduces jitter by 25--81\,\% across four models (MASt3R-SfM, VGGT, PI3,
PI3X) while preserving geometry to within $\pm$2--8\,\% Chamfer distance---a strictly
superior trade-off to every native 4D architecture and geometry-refinement baseline we
evaluate. At the two-camera extreme, VGGT combined with Opt beats the native 4D
architecture VGGT4D on every single metric.

%% -----------------------------------------------------------------------
\section{Background and Motivation}\label{s:background}

Classical multi-view reconstruction pipelines such as Structure-from-Motion
(SfM)~\cite{schoenberger2016sfm} produce accurate, metric-scale geometry in well-constrained
settings: they rely on repeatable keypoint matching and bundle adjustment, which deliver
precise camera poses and sparse point clouds when images share sufficient texture, overlap,
and baseline. In sparse-view configurations of 2--4 cameras, however, these conditions
frequently fail. Low inter-camera overlap means that few keypoints are visible in more than
one view simultaneously~\cite{hartley2004multiple}; short or degenerate baselines reduce
parallax and destabilise triangulation~\cite{schoenberger2016sfm}; and textureless
surfaces---common on human skin, clothing, and manipulation objects---yield too few reliable
correspondences for bundle adjustment to converge~\cite{dust3r, mast3r}. The result is
either outright reconstruction failure or incomplete, drift-prone geometry that cannot
support downstream tasks such as grasping or object tracking.

Learning-based spatial models such as DUSt3R~\cite{dust3r}, PI3, and PI3X~\cite{pi3}
overcome many of these failure modes by directly regressing dense 3D pointmaps from image
pairs, using learned priors to fill in low-texture and low-overlap regions where classical
methods collapse. These models achieve strong per-frame accuracy---PI3 and PI3X in
particular deliver the highest accuracy and completeness at a strict 1\,cm robotics
threshold among all models we evaluate. Although they address many of the correspondence
and reconstruction failures that limit classical SfM, they do not eliminate the broader
challenge of producing reconstructions that are sufficiently stable for downstream robotic
use. As purely spatial estimators, they process each frame independently and therefore
provide no explicit mechanism for enforcing temporal consistency across a sequence. The
resulting frame-to-frame instability can manifest as high-frequency jitter, scale
fluctuations, or pose inconsistencies, with the severity varying substantially across
architectures. Such instability can render otherwise accurate reconstructions unsuitable
for downstream tasks including trajectory estimation, object tracking, and robotic
manipulation.

GGPT~\cite{ggpt} augments feed-forward reconstructions using geometry-guided refinement informed by sparse SfM-derived correspondences. While the method improves reconstruction quality on the static benchmarks for which it was designed, its behaviour under the sparse-view dynamic conditions considered in this thesis differs substantially. Across all evaluated backbone models, GGPT consistently reduces the measured static-scene jitter metric, often by an order of magnitude. However, this reduction is accompanied by large increases in Chamfer distance and inconsistent changes in accuracy and completeness. Since the degradation is already evident in the per-frame evaluation prior to temporal alignment, the observed jitter suppression is unlikely to arise from improved geometric fidelity. Instead, the results suggest that GGPT acts as a strong geometric regulariser in this setting, producing more temporally stable but less geometrically accurate reconstructions.

The natural response is to incorporate temporal modelling directly. VGGT4D~\cite{vggt4d}
extends spatial transformers with motion cues, and MonoFusion~\cite{monofusion} fuses
monocular depth priors across time. Both enforce sequence-level consistency, but at a cost
we measure precisely: VGGT4D inflates Chamfer distance by 20--160\,\% relative to its
per-frame baseline, and MonoFusion suffers systematic alignment convergence failures in
sparse-view settings, producing duplicated subject artefacts rather than coherent
reconstructions.

None of these approaches simultaneously achieves geometric fidelity and temporal
consistency under sparse-view dynamic conditions. These results motivate the central design
principle of this work: temporal consistency and geometric fidelity are not in fundamental
conflict---they simply require a targeted rather than global smoothing strategy.

%% -----------------------------------------------------------------------
\section{Problem Statement and Contributions}\label{s:contributions}

We address the problem of temporally consistent, geometrically precise 4D reconstruction
from sparse multi-view video (2--4 cameras) of dynamic scenes, evaluated on the
DEX-YCB~\cite{dexycb} and Hi4D~\cite{hi4d} datasets. The main contributions of this work
are:

\begin{itemize}

  \item \textbf{Unified 4D Benchmarking Framework.}
        A structured evaluation pipeline that subjects six base models---MASt3R-SfM, VGGT,
        PI3, PI3X, VGGT4D, and MonoFusion---to identical data-loading, semantic masking, and
        metric-computation conditions across DEX-YCB~\cite{dexycb} and Hi4D~\cite{hi4d}.
        The framework measures both 3D per-frame and 4D sequence-level performance via
        Chamfer distance, accuracy and completeness at a 1\,cm threshold, temporal jitter,
        drift, and camera pose error, with static/dynamic decomposition where applicable
        (see \Cref{c:methodology}). It additionally evaluates the effect of GGPT~\cite{ggpt}
        geometry refinement applied on top of all base models except MonoFusion.

  \item \textbf{Comparative Temporal Alignment Study.}
        An analysis of three strategies for lifting per-frame 3D reconstructions into a
        globally consistent 4D representation (see \Cref{c:methodology}).
        The Reference strategy aligns every frame to a common anchor using static-scene
        correspondences and the Umeyama similarity transform.
        The Hierarchical strategy builds a binary tree of pairwise alignments, propagating
        consistency bottom-up.
        The Pose-Graph Optimisation strategy computes all $\mathcal{O}(T^2)$ pairwise edges,
        weights them by alignment reliability, and refines all frame poses simultaneously
        via robust geodesic averaging on $\mathrm{SO}(3)$.
        The Hierarchical strategy consistently delivers the best static-scene completeness
        and geometric stability; Pose-Graph Optimisation minimises the temporal consistency
        cost $\Delta_{\text{consistency}}$ (Equation~\ref{eq:delta_consistency}) and best
        preserves dynamic-region accuracy.

        \begin{equation}\label{eq:delta_consistency}
            \Delta_{\text{consistency}} = \mathrm{Chamfer}_{\mathrm{4D}}
                                        - \mathrm{Chamfer}_{\mathrm{3D}}
        \end{equation}

        This quantity isolates the spatial degradation introduced purely by coordinate
        alignment, enabling direct comparison of alignment strategies and model families.

  \item \textbf{Confidence-Weighted Temporal Smoothing (Opt).}
        A ground-truth-free post-processing method that reduces frame-to-frame instability
        for any per-frame estimator without retraining (see \Cref{c:refinement}).
        Opt applies a Gaussian temporal filter to already-aligned pointmaps, weighted by
        model confidence and restricted to pixels classified as static by
        SAM2~\cite{sam2}-derived masks---leaving dynamic pixels untouched to avoid
        suppressing real motion.
        A Pareto-frontier sweep over temporal bandwidth
        $\sigma \in \{0.0, 0.5, 1, 2, 4, 6, 8, 12, 20, 30\}$ and blending strength
        $\alpha \in \{0.0, 0.2, 0.5, 0.8, 1.0\}$ demonstrates that the optimal configuration is 
        model-dependent: MASt3R-SfM benefits from a wider temporal window
        ($\sigma{=}6$, $\alpha{=}1.0$), PI3 and PI3X from a moderate window
        ($\sigma{=}4$, $\alpha{=}1.0$), and VGGT from a narrower window
        ($\sigma{=}2$, $\alpha{=}1.0$), reflecting distinct jitter profiles across
        architectures. These per-model configurations achieve a 25--81\,\% jitter
        reduction while preserving geometry to within $\pm$2--8\,\% Chamfer distance---a
        strictly superior trade-off to every native 4D architecture and
        geometry-refinement baseline evaluated in this thesis.

\item \textbf{Empirical Characterisation of Architectural Families.}
      A systematic comparison of spatial-first models (PI3, PI3X, MASt3R-SfM, VGGT),
      the geometry-refinement approach GGPT, and native 4D architectures (VGGT4D,
      MonoFusion) reveals distinct and quantifiable failure modes under sparse-view
      dynamic conditions. PI3 and PI3X consistently achieve the strongest geometric
      performance across both DEX-YCB and Hi4D. MASt3R-SfM exhibits the greatest
      frame-to-frame instability, making it the model that benefits most from temporal
      refinement. VGGT provides a competitive balance between accuracy and stability.
      GGPT worsens Chamfer distance by 3--10$\times$ and produces inconsistent accuracy
      and completeness behaviour across architectures and view counts, while suppressing
      jitter to near-zero---a combination that makes it unsuitable for geometry-critical
      applications despite its apparent temporal stability. Native 4D architectures
      reduce temporal noise but consistently sacrifice geometric accuracy relative to
      spatial-first models. Together these findings establish the decoupled
      framework---strong spatial estimation, robust temporal alignment, and
      confidence-weighted temporal smoothing---as the most effective approach for
      sparse-view 4D reconstruction.

\end{itemize}