\chapter{Literature Review and Background}
\label{c:background}

This chapter reviews the theoretical and technical foundations underlying the reconstruction pipelines evaluated throughout this thesis. Section~\ref{sec:sfm_mvs_foundations} introduces classical Structure-from-Motion (SfM) and Multi-View Stereo (MVS) methods, highlighting both their strengths and their limitations under sparse-view conditions. Section~\ref{sec:foundation_models} surveys recent foundation models for 3D vision, grouped according to their treatment of spatial and temporal information. Section~\ref{sec:sam2_background} describes the Segment Anything Model 2 (SAM2), which is used throughout this work to distinguish static and dynamic scene regions. Finally, Section~\ref{sec:math_background} presents the mathematical concepts required for the proposed alignment, including similarity transformations, rotation averaging on $\mathrm{SO}(3)$, robust geodesic averaging, and geometric evaluation metrics.

\section{Structure-from-Motion and Multi-View Stereo Foundations}
\label{sec:sfm_mvs_foundations}

Three-dimensional scene reconstruction from multiple images has traditionally been addressed through the complementary paradigms of Structure-from-Motion (SfM)~\cite{hartley2004multiple,schoenberger2016sfm} and Multi-View Stereo (MVS)~\cite{schoenberger2016mvs}. Together, these methods form the foundation of modern photogrammetry systems and remain the dominant approach for producing accurate metric-scale reconstructions from calibrated or partially calibrated image collections.

SfM seeks to jointly estimate camera poses and scene structure from image observations. Given a collection of overlapping images, distinctive visual features are first detected and matched across views. Geometric verification is then used to reject inconsistent correspondences, after which camera poses and sparse three-dimensional points are estimated through triangulation and refined using bundle adjustment~\cite{hartley2004multiple,schoenberger2016sfm}. Established SfM systems such as COLMAP~\cite{schoenberger2016sfm} achieve remarkable accuracy and robustness on large-scale reconstruction tasks, often producing camera trajectories and sparse point clouds with low reprojection error. More recent systems such as GLOMAP~\cite{pan2024glomap} and MP-SfM~\cite{pataki2025mpsfmmonocularsurfacepriors} have pushed reconstruction quality further through global pose initialisation and monocular surface priors respectively.

While SfM recovers only a sparse representation of scene geometry, MVS extends this sparse reconstruction into a dense surface estimate. Given camera poses obtained from SfM, MVS algorithms estimate depth for individual pixels by enforcing photometric consistency across multiple views. These depth estimates are subsequently fused into dense point clouds or surface representations~\cite{schoenberger2016mvs,seitz2006comparison}. Modern MVS systems employ sophisticated view-selection strategies and regularisation techniques that enable high-quality reconstructions across a wide range of scenes~\cite{schoenberger2016mvs}.

Despite their success, classical SfM and MVS pipelines rely fundamentally on the existence of reliable multi-view correspondences. Performance degrades substantially when these assumptions are violated. Sparse-view reconstruction represents a particularly challenging regime because the number of observations available for establishing geometric constraints is severely limited. When only two to four cameras are available, several failure modes become prominent.
First, low inter-camera overlap reduces the number of features that are simultaneously visible across views, limiting the quantity and quality of geometric correspondences available for pose estimation~\cite{hartley2004multiple,pataki2025mpsfmmonocularsurfacepriors}. Second, short or degenerate baselines reduce triangulation accuracy because depth uncertainty grows rapidly as the triangulation angle decreases~\cite{hartley2004multiple,schoenberger2016sfm}. Third, textureless surfaces such as human skin, clothing, walls, and many manipulation objects provide few distinctive features for matching, causing correspondence estimation and bundle adjustment to become unstable~\cite{dust3r,mast3r}. Fourth, and most critically for the setting considered in this thesis, dynamic scene content introduces a qualitatively different challenge: moving objects actively violate the static-world assumption underlying both feature matching and bundle adjustment, producing spurious correspondences that cannot be resolved by increasing image count or improving texture~\cite{Cin_2025_CVPR}. Unlike the first three failure modes, which can in principle be mitigated by careful rig design or denser capture, dynamic content represents a fundamental incompatibility with the geometric model assumed by classical pipelines.

These limitations are particularly acute in robotics and human-centred perception scenarios, where sparse multi-camera systems are often preferred for reasons of cost, deployment complexity, and occlusion constraints, yet such configurations provide exactly the conditions under which traditional SfM and MVS become least reliable. The resulting reconstructions may be incomplete, geometrically inconsistent, or entirely unsuccessful, limiting their usefulness for downstream tasks such as object tracking, trajectory estimation, manipulation planning, and scene understanding.

\section{Foundation Models for 3D and 4D Reconstruction}
\label{sec:foundation_models}

Rather than relying on geometric matching alone, modern foundation models learn strong scene priors from large-scale image collections, enabling them to recover plausible geometry even in regions with limited texture, weak overlap, or ambiguous correspondence structure~\cite{dust3r,mast3r,vggt}. Although these methods share a common goal of reconstructing scene structure from images, they differ substantially in how geometry and time are represented.

The models evaluated in this thesis are grouped into three architectural families according to their primary design objective. \textit{Spatial reconstruction models} estimate scene geometry from a collection of images without explicit temporal modelling (MASt3R-SfM, VGGT, Pi3, and Pi3X). \textit{Geometry refinement models} improve existing reconstructions through additional geometric reasoning (GGPT). \textit{Temporal reconstruction models} jointly model geometry and motion to produce temporally coherent reconstructions (VGGT4D and MonoFusion). Together, these families represent the principal design directions currently being explored in modern 3D and 4D vision, and their differing assumptions about geometry, correspondence, and temporal consistency form the basis for the comparative evaluation presented in Chapters~\ref{c:methodology} and~\ref{c:results}.

\subsection{Spatial Reconstruction Models}
\label{sec:spatial_models}

Spatial reconstruction models focus on recovering accurate scene geometry from a set of observations without explicitly modelling temporal evolution. Their primary objective is geometric reconstruction, with camera poses, depth, or pointmaps inferred directly from one or more images using learned geometric priors. While these models can be applied independently to each frame of a video sequence, they provide no mechanism for enforcing temporal consistency across successive reconstructions.

\paragraph{MASt3R-SfM.}
MASt3R-SfM~\cite{mast3rsfm} is a hybrid reconstruction pipeline that combines learned
geometric reasoning with classical SfM optimisation. The approach builds upon
DUSt3R~\cite{dust3r}, which introduced dense pointmap regression as an alternative to
traditional depth estimation and triangulation, and MASt3R~\cite{mast3r}, which further
improved robustness by grounding image matching directly in predicted three-dimensional
geometry. Both DUSt3R and MASt3R operate on image pairs: given two views, the model
regresses a dense pointmap for each, expressed in the coordinate frame of the first.
MASt3R-SfM integrates these pairwise predictions into a complete SfM pipeline, using
geometry-aware correspondences to initialise camera poses and scene structure before
applying global bundle adjustment. By combining learned pairwise matching with classical
global optimisation, the system achieved strong performance on challenging unconstrained
reconstruction benchmarks while retaining the geometric consistency advantages of
classical SfM.

\paragraph{VGGT.}
VGGT~\cite{vggt} (Visual Geometry Grounded Transformer) moves beyond pairwise reasoning
by processing all input views jointly within a single forward pass. Rather than
accumulating pairwise predictions through a subsequent optimisation stage, VGGT uses a
transformer architecture trained to recover cameras, depth, and scene structure from
multiple images simultaneously, learning global geometric relationships directly from
large-scale data. This design reduces dependence on handcrafted reconstruction stages
and achieves strong performance across a wide variety of reconstruction tasks. Its
architecture reflects a broader trend toward general-purpose geometric foundation models
capable of solving multiple three-dimensional vision problems within a single learned
representation.
\paragraph{Pi3 and Pi3X.}

Pi3~\cite{pi3} introduces a reference-free formulation of visual geometry reconstruction that removes the need for a designated anchor image. Using a fully permutation-equivariant architecture, the model jointly predicts camera poses and dense point maps from an unordered set of images, making reconstruction inherently robust to input ordering and reference-view selection. By learning geometric priors from large-scale data, Pi3 can recover accurate scene structure even in sparse-view settings where traditional correspondence-based methods become unreliable. Pi3X extends this framework with improved point-map decoding, more reliable
confidence estimation, optional conditioning on camera and depth priors, and
support for approximate metric-scale reconstruction. These modifications are
designed to improve reconstruction quality while preserving the reference-free
geometric reasoning of the original model, though their effect on Chamfer
distance under sparse-view conditions is examined in Chapter~\ref{c:results}.

\subsection{Geometry Refinement Models}
\label{sec:refinement_models}

A complementary class of methods seeks to improve reconstruction quality without replacing the underlying reconstruction backbone. Instead of generating geometry directly, these approaches operate as post-processing stages that refine an existing geometric prediction using additional evidence.

\paragraph{GGPT.}

GGPT~\cite{ggpt} (Geometry-Grounded Point Transformer) is a geometry-refinement framework designed to enhance feed-forward three-dimensional reconstructions through geometry-guided correction. The method combines dense pointmap predictions with incomplete geometric cues derived from an auxiliary SfM process, using a transformer-based refinement network to improve consistency and reconstruction quality. Conceptually, GGPT occupies a middle ground between classical optimisation-based reconstruction and purely feed-forward prediction: geometric structure is first estimated by a backbone model and subsequently refined using additional geometric evidence.

Unlike spatial reconstruction and temporal reconstruction architectures, GGPT does not define a complete reconstruction pipeline in its own right. Instead, it functions as a model-agnostic refinement stage that can be applied to the outputs of existing reconstruction systems. By incorporating additional geometric constraints during refinement, GGPT aims to improve the consistency and quality of an existing reconstruction rather than generating geometry directly. Such refinement may also affect the temporal behaviour of reconstructed sequences, since stronger geometric regularisation can suppress frame-to-frame variations in the predicted geometry. This design makes GGPT particularly relevant for understanding whether reconstruction quality is best improved through stronger backbone architectures, explicit temporal modelling, or post-hoc geometric refinement.

\subsection{Temporal Reconstruction Models}
\label{sec:temporal_models}

A common limitation of purely spatial reconstruction models is the absence of explicit temporal reasoning: each frame or image set is reconstructed independently, meaning that geometric predictions may vary between consecutive frames even when the underlying scene remains static. Temporal reconstruction models address this by incorporating motion information and sequence-level reasoning during inference, jointly modelling geometry and temporal evolution rather than treating each frame as an isolated input~\cite{vggt4d}.

\paragraph{VGGT4D.}

VGGT4D~\cite{vggt4d} extends the VGGT architecture to four-dimensional reconstruction by incorporating temporal motion cues into the transformer backbone. The model is designed to reason jointly about spatial structure and temporal evolution, enabling reconstruction of dynamic scenes across time. By leveraging information from neighbouring frames, VGGT4D aims to improve temporal consistency and reduce frame-to-frame fluctuations that commonly arise in per-frame reconstruction systems.

\paragraph{MonoFusion.}

MonoFusion~\cite{monofusion} approaches sparse-view four-dimensional reconstruction from a different perspective. Instead of relying solely on multi-view geometric constraints, it combines monocular depth priors with temporal fusion mechanisms to accumulate information across a sequence. Concretely, the model aligns per-camera monocular depth estimates to DUSt3R-predicted multi-view consistent pointmaps, which serve as a metric-scale reference for the iterative temporal fusion. This alignment is performed iteratively across views and time to produce a temporally coherent reconstruction. The model reflects a growing trend toward integrating learned monocular geometry with multi-view reasoning to overcome the limitations of sparse observations~\cite{li2025mono3rexploitingmonocularcues,Jiang_2025_ICCV,Cheng2025MonSterUS}.

While temporal architectures address an important limitation of spatial models by introducing sequence-level consistency, the incorporation of temporal information inevitably introduces additional modelling assumptions and complexity. This creates a trade-off between geometric precision and temporal stability that remains an active area of research, and one that the comparative evaluation in this thesis aims to illuminate under controlled sparse-view conditions on two dynamic multi-view benchmarks.

\section{SAM2 for Static and Dynamic Scene Segmentation}
\label{sec:sam2_background}

SAM2~\cite{sam2} (Segment Anything Model 2) extends the original Segment Anything Model from static-image segmentation to image and video sequences. By combining promptable segmentation with a memory mechanism that propagates object information across time, SAM2 produces temporally consistent masks throughout a video sequence. Trained on large-scale image and video datasets, the model demonstrates strong zero-shot generalisation across diverse object categories and scene types.

Video segmentation has become increasingly relevant for dynamic-scene reconstruction, where moving actors and objects violate the static-world assumptions commonly exploited by geometric reconstruction methods. Recent systems such as MonoFusion~\cite{monofusion} employ foreground-background segmentation to distinguish dynamic scene content from temporally stable geometry during reconstruction and optimisation.

In this thesis, SAM2 is used to generate per-frame static and dynamic segmentation masks. These masks serve as a shared input to several downstream components: they enable separate evaluation of reconstruction quality in static and dynamic regions, provide the static-scene correspondences required by the temporal alignment strategies, and gate the confidence-weighted temporal smoothing filter so that only static-region predictions are smoothed across time. The specific procedures for mask generation and integration are described in Chapter~\ref{c:methodology}.

\section{Mathematical Background}
\label{sec:math_background}

The reconstruction models evaluated in this thesis do not necessarily predict geometry in a common coordinate system. In particular, the spatial foundation models discussed in Section~\ref{sec:foundation_models} reconstruct each frame or image set independently, often producing point clouds that differ by an arbitrary scale, rotation, and translation. Constructing a temporally coherent 4D reconstruction therefore requires estimating geometric transformations between independently reconstructed frames and combining these estimates into a globally consistent coordinate system.

This section introduces the mathematical concepts that underpin these operations. First, similarity transformations and Umeyama alignment are presented as the basis for point-cloud registration. Next, the geometry of the rotation manifold $\mathrm{SO}(3)$ and geodesic rotation averaging are introduced, which form the foundation of the alignment graph optimisation strategy described later in Chapter~\ref{c:methodology}. Finally, the Chamfer distance is presented as a primary metric for evaluating reconstruction quality.

\subsection{Similarity Transformations and Umeyama Alignment}

A similarity transformation accounts for the arbitrary scale, rotation, and
translation that may separate two independently reconstructed point clouds:
\begin{equation}
\mathbf{y} = s\,\mathbf{R}\mathbf{x} + \mathbf{t},
\end{equation}
where $s \in \mathbb{R}^{+}$ denotes a global scale factor,
$\mathbf{R} \in \mathrm{SO}(3)$ is a rotation matrix, and
$\mathbf{t} \in \mathbb{R}^{3}$ is a translation vector.
Throughout this thesis, similarity alignment is performed using the Umeyama
algorithm~\cite{umeyama1991least}, a closed-form least-squares method that
estimates the optimal scale, rotation, and translation minimising the mean
squared alignment error over a set of corresponding three-dimensional points.
Because the solution is available in closed form, it provides an efficient and
numerically stable primitive for registering reconstructed point clouds.
The estimated similarity transformations serve two purposes: aligning
reconstructions to ground truth during evaluation, and computing frame-to-frame
alignments within the temporal reconstruction strategies introduced in
Chapter~\ref{c:methodology}.
\subsection{Rotation Geometry on \texorpdfstring{$\mathrm{SO}(3)$}{SO(3)}}

While translations and scales can be manipulated using standard Euclidean operations, rotations require special treatment because they lie on a curved manifold rather than in a vector space. The set of valid three-dimensional rotations forms the special orthogonal group
\begin{equation}
\mathrm{SO}(3) = \left\{ \mathbf{R}\in\mathbb{R}^{3\times3} \;\middle|\; \mathbf{R}^{\top}\mathbf{R}=\mathbf{I},\; \det(\mathbf{R})=1 \right\}.
\end{equation}

Direct arithmetic averaging of rotation matrices generally produces matrices that no longer satisfy the orthogonality constraints of $\mathrm{SO}(3)$~\cite{absil2008optimization,chirikjian2011stochastic}. Consequently, optimisation and averaging operations must be performed using geometric constructions defined on the manifold itself.

A commonly used measure of rotational discrepancy is the geodesic distance
\begin{equation}
d(\mathbf{R}_1,\mathbf{R}_2) = \arccos\!\left(\frac{\mathrm{tr}\!\left(\mathbf{R}_1^{\top}\mathbf{R}_2\right)-1}{2}\right),
\end{equation}
which corresponds to the minimum rotation angle required to transform one orientation into the other~\cite{hartley2013rotation}.

\subsection{Geodesic Rotation Averaging}

The alignment graph optimisation strategy developed in this thesis estimates pairwise similarity transformations between many frames in a sequence. Since multiple alignment paths may exist between two frames, individual rotation estimates must be combined into a single globally consistent orientation.

Given a set of rotations $\{\mathbf{R}_i\}_{i=1}^{N}$, rotation averaging seeks a consensus rotation that best represents the collection. A common formulation minimises the sum of squared geodesic distances,
\begin{equation}
\mathbf{R}^{*} = \arg\min_{\mathbf{R}\in\mathrm{SO}(3)} \sum_{i=1}^{N} d(\mathbf{R},\mathbf{R}_i)^2.
\end{equation}

Although effective, this L2 formulation can be sensitive to erroneous pairwise alignments and outliers. To improve robustness, this thesis employs an L1 geodesic formulation,
\begin{equation}
\mathbf{R}^{*} = \arg\min_{\mathbf{R}\in\mathrm{SO}(3)} \sum_{i=1}^{N} d(\mathbf{R},\mathbf{R}_i),
\end{equation}
which corresponds to the geometric median on the rotation manifold.

The optimisation is performed using a manifold adaptation of the Weiszfeld algorithm~\cite{weiszfeld1937point}, an iteratively reweighted least-squares procedure in which each input is assigned a weight inversely proportional to its distance from the current estimate, causing outliers to receive progressively less influence. On the rotation manifold, the algorithm operates by mapping the current estimate and all input rotations into the local tangent space via the logarithmic map $\log: \mathrm{SO}(3) \to \mathfrak{so}(3)$---the space of skew-symmetric matrices that captures small rotational displacements---computing a distance-weighted mean in that linear space, and projecting the result back onto $\mathrm{SO}(3)$ via the exponential map $\exp: \mathfrak{so}(3) \to \mathrm{SO}(3)$.

\subsection{Chamfer Distance}
\label{sec:chamfer_distance}

Geometric reconstruction quality is evaluated throughout this thesis using a combination of distance-based and threshold-based metrics. The primary distance-based metric is the symmetric Chamfer distance, which measures the average geometric discrepancy between reconstructed and ground-truth point clouds~\cite{chamfer1977image}.

Given two point sets $P$ and $Q$, the Chamfer distance is defined as
\begin{equation}
d_{\mathrm{Chamfer}}(P,Q) = \frac{1}{|P|}\sum_{\mathbf{p}\in P}\min_{\mathbf{q}\in Q}\|\mathbf{p}-\mathbf{q}\|_2 + \frac{1}{|Q|}\sum_{\mathbf{q}\in Q}\min_{\mathbf{p}\in P}\|\mathbf{q}-\mathbf{p}\|_2.
\end{equation}

It is worth noting that the two terms in this definition are not symmetric in meaning. The first term measures \emph{accuracy}: how well each reconstructed point finds a close match in the ground truth. The second term measures \emph{completeness}: how much of the ground truth geometry is covered by the reconstruction. Smaller values therefore indicate closer geometric agreement between the two point sets.