\chapter{Discussion and Conclusion}
\label{c:discussion}

This thesis set out to answer a single question: can sparse-view dynamic reconstruction
achieve both geometric fidelity and temporal consistency simultaneously, without
sacrificing one for the other? The results of Chapter~\ref{c:results} provide a clear
answer. The decoupled paradigm of a strong per-frame spatial estimator, training-free
temporal alignment, and targeted static-region smoothing consistently achieves both
properties across every backbone and every view count evaluated. No existing alternative
matches this trade-off. This chapter interprets why, identifies what remains unsolved,
and proposes directions for future work.

%% -----------------------------------------------------------------------
\section{Architecture Family Analysis}
\label{sec:architecture_analysis}

The central finding of Chapter~\ref{c:results} is that native 4D architectures fail to
close the geometry--jitter trade-off that spatial models with Opt smoothing achieve,
which the following sections explain by examining each family's architectural assumptions.

\subsection{Why Spatial Models Succeed}
Spatial foundation models reconstruct each frame independently, using learned priors to
recover dense geometry from sparse observations. Independence is a strength: each frame
is solved without any assumption about what neighbouring frames look like, so
reconstruction errors do not propagate across time. Pi3 achieves the highest geometric fidelity among all evaluated models, reaching
0.019\,m Chamfer and 95.9\,\% static accuracy at four views on DexYCB, precisely
because no temporal constraint forces it to compromise geometric detail for sequence
consistency.

\subsection{Why Native 4D Architectures Fail}
The fundamental limitation of native 4D architectures and temporal estimators stems from a structural coupling: they embed temporal consistency directly into the core scene reconstruction loop, sacrificing frame-level geometric fidelity to enforce sequence-level smoothness~\cite{vggt4d, monofusion}. 

VGGT4D, on the one hand, processes the full multi-view video jointly and enforces consistency through a shared transformer representation~\cite{vggt4d}. This joint processing imposes a steep geometric cost, elevating Chamfer distances relative to independent per-frame backbones. Although VGGT4D explicitly attempts to decouple static and dynamic regions through attention-based dynamic cue mining, the qualitative analysis in Section~\ref{sec:qualitative} demonstrates that this decoupling remains incomplete under sparse-view conditions. Under these constrained views, the joint architecture produces severe duplication artefacts and structural smearing. These errors stem directly from the temporal averaging of conflicting depth hypotheses. When the joint architecture receives a new viewpoint whose depth estimates conflict with existing ones, the model averages the hypotheses rather than resolving them. This averaging displaces the reconstructed surface and significantly reduces completeness, explaining the non-monotonic completeness patterns observed under sparse viewpoints. 

MonoFusion, on the other hand, illustrates how optimization-based temporal estimators fall into the same structural trap. Instead of maintaining a shared internal representation, its iterative monocular-to-multi-view depth alignment loop relies on global affine constraints calculated strictly from background pixels~\cite{monofusion}. This specific heuristic remains highly susceptible to local optima under wide camera spacing and large subject-to-camera distances. Consequently, MonoFusion produces duplicated-subject artefacts rather than coherent reconstructions. While this alignment strategy minimizes drift and balances mean jitter, its frame-level geometric fidelity remains the weakest among all evaluated models.

Ultimately, these temporal estimators do not provide a principled solution to the geometry--jitter trade-off under sparse-view conditions; they merely trade one structural failure mode for another.

\subsection{Why GGPT Fails Under Sparse-View Dynamic Conditions}
As documented in Section~\ref{sec:native4d_baselines}, GGPT substantially degrades
geometry across every backbone and both datasets under the sparse-view
conditions evaluated in this thesis. Like all methods evaluated here, GGPT
operates partially out of domain: it is trained on ScanNet++ with 4--12 input
views per sequence~\cite{ggpt}, whereas this thesis evaluates two to four views
of dynamic scenes. However, while spatial foundation models generalise
reasonably under these conditions despite their own domain gap, GGPT's
refinement mechanism has a specific dependency that spatial models do not:
it relies on SfM-derived geometric cues to ground its transformer network.
Under sparse-view dynamic conditions, these auxiliary cues are unreliable,
and grounding a refinement network on unreliable geometric evidence
evidently introduces systematic distortion rather than correcting it.

%% -----------------------------------------------------------------------
\section{The decoupled Framework Paradigm}
\label{sec:decoupled_paradigm}

The results of Chapter~\ref{c:results} provide consistent evidence for a design
principle: separating the problem of geometric reconstruction from the problem of
temporal consistency, and solving each with the right tool, outperforms joint
end-to-end approaches.

The argument has three parts. First, spatial foundation models achieve higher geometric
fidelity than native 4D architectures at the per-frame level, before any temporal
processing is applied; as established in Section~\ref{sec:native4d_baselines}, Pi3
outperforms VGGT4D by 63--72\,\% on DexYCB Chamfer distance across two- and four-view
configurations (0.018--0.019\,m vs.\ 0.049--0.069\,m). Second, temporal consistency can be added after the
fact without destroying geometry, provided the smoothing is applied only where it is
valid, namely static regions. Opt reduces mean jitter by 60--91\,\% across four backbones and eliminates
high-frequency jitter entirely (Section~\ref{sec:native4d_baselines}), while leaving
Chamfer flat or improved for all but Pi3X. Third, the
resulting pipeline Pareto-dominates native 4D architectures on the geometry--jitter
trade-off at every view count for every backbone evaluated. Pi3+Opt achieves 0.025\,m
Chamfer and 0.001\,m mean jitter at four views; VGGT4D achieves 0.069\,m Chamfer and
0.003\,m jitter at the same view count.

The key insight is that geometric accuracy and temporal consistency conflict only when
temporal constraints are applied globally, penalising dynamic regions where motion is
real and should be preserved. The decoupled paradigm avoids this conflict by design:
the spatial estimator is free to maximise geometric fidelity without any temporal
pressure, and Opt is free to maximise temporal consistency without any geometric
pressure, because it operates exclusively on static regions where consistency is
physically guaranteed.

The choice of alignment strategy follows the same decoupling logic at a smaller scale,
with Strategy 2 selected over both alternatives for different reasons in each case.
Against Strategy 1, robustness is the deciding factor. Strategy 1 anchors every frame
directly to a single reference frame $t_0$, so alignment accuracy for any frame depends
entirely on how well that frame is observed and how far the scene has moved from its
initial configuration (Section~\ref{sec:strategy1}). Strategy 2 removes this single
point of failure by merging frames hierarchically, so no frame's alignment depends on
one reference that could be poorly observed. Table~\ref{tab:strategy_extended} shows no
consistent empirical advantage for Strategy 2 over Strategy 1 on DexYCB, since both
achieve comparable Chamfer and $\Delta_{\text{consistency}}$ values for every backbone;
the choice rests on robustness to a reference-frame failure that the bounded DexYCB
sequences happen not to exhibit, not on an observed gap. A longer sequence, or one with
a poorly observed first frame, would be expected to widen the difference.

The same robustness argument favours Strategy 3 over Strategy 2 under harder
conditions than those evaluated here. Coupling every frame to every other frame through
$\mathcal{O}(T^2)$ reliability-weighted edges lets Strategy 3 down-weight a single
unreliable pairwise alignment rather than propagate it, unlike Strategy 2's hierarchical
merge, where a poor alignment at one tree level corrupts every frame downstream. Longer
sequences, where the merge tree has more levels for an error to propagate through, or
sequences with intermittently unreliable frames, such as motion blur or partial
occlusion at specific timesteps, are where this advantage should matter most: Strategy
3's edge-reliability weighting can isolate the affected frames rather than letting them
corrupt their neighbours in the merge tree. A fixed 50-iteration cap on the alignment
graph optimisation (Section~\ref{sec:strategy3}) further understates Strategy 3's true
robustness in the results reported here, since a harder alignment problem could in
principle require more iterations than the budget allows. None of these harder
conditions are pronounced enough under the bounded, 24--25 frame DexYCB and Hi4D
sequences evaluated in this thesis for the advantage to materialise: Strategy 3 achieves
marginally better $\Delta_{\text{consistency}}$ in most cases, but the differences
between strategies are modest throughout and lie within one standard deviation across
all model families on DexYCB. Computational cost therefore drives the final choice.
Strategy 3's $\mathcal{O}(T^2)$ pairwise alignments and iterative geodesic averaging make
it substantially slower without a proportionate benefit under these conditions, while
Strategy 2's $\mathcal{O}(T \log T)$ complexity and consistent performance across all
four backbones make it the more suitable foundation for a post-processing pipeline
intended for practical deployment. Strategy 3 should remain the preferred choice for
longer or less reliable sequences in future work.
%% -----------------------------------------------------------------------
\section{Limitations}
\label{sec:limitations}

Five limitations of the proposed approach warrant explicit discussion.

\subsection{Per-Backbone Parameter Calibration}
As discussed in Section~\ref{sec:opt_parameters}, Opt requires per-backbone
values of $\sigma$ and $\alpha$, and the selected configurations represent
single operating points on the Pareto frontier of each backbone. More
fundamentally, the jitter level produced by a spatial backbone depends on the
reconstruction confidence of that backbone in the current scene, which varies
with scene type, texture, lighting, and the distribution of the backbone's
training data. A parameter set calibrated on DexYCB manipulation sequences
may not transfer to surgical, outdoor, or industrial scenes without
recalibration.

\subsection{Drift is Not Corrected}
Opt suppresses high-frequency jitter and reduces mean jitter substantially, but
does not reliably reduce drift. Drift is a slow, monotonic accumulation of pose
estimation errors that displaces static anchors over the sequence. Because the
Gaussian filter operates symmetrically in time, it attenuates rapid oscillations
effectively but cannot correct a bias that accumulates monotonically. The
increase observed for Pi3 at two views, where drift rises from 0.007\,m under
Strategy 2 to 0.034\,m under Opt, reflects the unusually low baseline drift of
Strategy 2 for this backbone: temporal smoothing redistributes pointmap positions
along the sequence, which can introduce long-term displacement where little
existed before. Correcting drift requires a mechanism that operates globally
across the sequence rather than locally in time, such as the anchor-constrained
optimisation described in Section~\ref{sec:future_work}.

\subsection{Static Mask Quality}
The SAM2-derived static masks are a critical dependency of both the alignment
strategies and Opt. Mask quality degrades in two scenarios: motion that is too
subtle to produce a strong frame-difference prior, leaving SAM2 with insufficient
signal to identify the moving subject, and scene boundaries where static and
dynamic regions are spatially adjacent. Boundary pixels at the static-dynamic
boundary are incidentally affected by Opt smoothing: projecting smoothed boundary
points back into 2D shifts their apparent class assignment, producing a reduction
in reported dynamic accuracy beyond the direct effect of the filter. This boundary
reclassification effect is most pronounced for Pi3X, which achieves the highest
baseline dynamic accuracy of all evaluated backbones; a higher proportion of
correctly classified dynamic boundary pixels means the reclassification effect
has more to act on, producing a larger absolute drop in reported dynamic accuracy
under Opt.

\subsection{Evaluation Scope}
Opt is evaluated exclusively on DexYCB, since Hi4D provides no static background
against which jitter metrics can be computed. The temporal consistency results
therefore rest on a single dataset covering human--object manipulation under controlled
laboratory conditions. Whether the 60--91\,\% jitter reductions reported here
generalise to outdoor scenes, surgical environments, or datasets with different camera
geometries and motion profiles remains an open question. The alignment strategy
comparison covers both datasets, but Hi4D results reflect the performance of the
underlying models rather than the properties of the strategies, since static
correspondences are absent after preprocessing.

\subsection{Confidence Threshold as a Global Compromise}
The 50\,\% confidence threshold used throughout Chapter~\ref{c:results}
(Section~\ref{sec:evaluation_pipeline}) was validated through a sweep
across all four spatial backbones (Figure~\ref{fig:conf_sweep}). This sweep
revealed that MASt3R-SfM, Pi3, and Pi3X share a similarly shaped Chamfer curve
with an interior minimum between 50\,\% and 80\,\%, while VGGT instead shows a
monotonic accuracy--completeness trade-off with no comparable interior optimum.
A single global threshold is therefore necessarily a compromise: it cannot be
simultaneously optimal for a backbone with a U-shaped response and a backbone
with a monotonic one. The 50\,\% threshold adopted in this thesis lies within
or close to the empirical optimum for three of the four backbones and avoids
the degenerate extremes for the fourth, but a per-backbone threshold, calibrated
in the same spirit as the per-backbone $\sigma$ and $\alpha$ values of
Section~\ref{sec:opt_parameters}, would likely yield a modest further
improvement in reported geometric fidelity. This limitation is methodological
rather than substantive: confidence filtering exists to prevent unreliable
correspondences from biasing the Umeyama alignment, not to optimise the
reported metrics directly, and an alternative such as confidence-weighted or
robust alignment (e.g., iterative closest point with outlier rejection~\cite{121791})
could remove the need for a hard threshold altogether.

%% -----------------------------------------------------------------------
\section{Future Work}
\label{sec:future_work}

This thesis established a foundation: spatial foundation models can be lifted to
temporally consistent 4D reconstruction through training-free alignment and a
training-free temporal low-pass filter, without sacrificing geometric fidelity. The
remaining open problems are clear. Drift is not corrected. The 2-view to 4-view
performance gap is not closed. Evaluation is limited to fixed, calibrated camera rigs
on two datasets. Future work can extend this foundation in four directions: improving
the smoothing method, correcting drift, broadening evaluation to moving cameras
and diverse scenes, and exploring learned refinement within the decoupled paradigm.

\subsection{Inference-Time Adaptive Parameter Selection}
The per-backbone parameter calibration limitation motivates replacing the offline sweep
with an adaptive procedure that selects $\sigma$ and $\alpha$ at inference time from
the current sequence. The required signal is available without ground truth: the
per-frame jitter of static anchors, computable from the aligned pointmaps before Opt
is applied, directly reflects the instability that $\sigma$ must suppress. A wider
$\sigma$ is appropriate when anchor jitter is high; a narrower $\sigma$ when it is
stable. Fitting $\sigma$ to the observed jitter profile of each sequence would remove
the dataset-specific calibration requirement and produce a method that generalises
without recalibration.

\subsection{Global Anchor-Constrained Drift Correction}
The drift limitation motivates a second post-processing stage applied after Opt.
After Opt suppresses jitter, static scene anchors are nearly stable across frames,
making their world-space positions reliably identifiable. The current alignment
strategies minimise pairwise alignment error sequentially, which allows drift to
accumulate along the propagation path. Treating static anchor positions as fixed
landmarks and solving simultaneously for all per-frame similarity transforms would
break this sequential structure: every frame is coupled to every anchor at once,
leaving no path along which errors can propagate monotonically. This direction is
structurally analogous to classical bundle adjustment applied to already-reconstructed
pointmaps rather than raw image correspondences, and requires no additional training
data.

\subsection{Reprojection-Error Cross-View Refinement}
A promising direction for future work is a refinement stage that targets the
cross-view geometric inconsistency driving the 2-view performance gap identified
in Section~\ref{sec:sparse_view}. Spatial foundation models are not trained to
produce outputs that agree across views; a post-processing stage that penalises
reprojection error using known camera intrinsics could close this gap without
additional cameras or retraining. Unlike temporal smoothing, reprojection-based
refinement would not require a static-scene assumption and could apply equally
to static and dynamic scene regions, making it a natural complement to the
proposed pipeline.

\subsection{Extension to Moving Cameras and Broader Evaluation}
This thesis evaluates exclusively on fixed, pre-calibrated camera rigs. Two datasets
motivate extending the evaluation framework to moving cameras. MultiCamRobolab~\cite{multicamrobolab}
introduces a real-world multi-camera dataset with ground-truth poses from a motion
capture system, targeting the setting of multiple freely moving cameras capturing a
shared dynamic scene. Extending the pipeline to this setting would require integrating
odometry data into the alignment stage, since the static mask derivation currently uses
pixel differences between frames and cannot distinguish camera-induced optical flow from
genuine scene motion. DyCheck~\cite{dycheck} provides a complementary evaluation
protocol for dynamic view synthesis from monocular video, introducing co-visibility
masked metrics and correspondence accuracy that overcome the multi-view signal leakage
present in existing benchmarks. Evaluating on both datasets would establish whether the
decoupled paradigm generalises beyond the controlled laboratory conditions of DexYCB
and Hi4D, and identify which components require adaptation for moving-camera or
single-camera settings.

%% -----------------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}

This thesis addressed the problem of temporally consistent 4D reconstruction from
sparse multi-view video. The core challenge, that existing methods achieve either
geometric fidelity or temporal consistency but not both, is resolved by a decoupled
paradigm that keeps the two objectives separate and solves each with the appropriate
tool.

Three contributions support this conclusion. A unified benchmarking framework evaluates
six reconstruction models under identical conditions across DexYCB and Hi4D, yielding
the first systematic comparison of spatial, geometry-refinement, and native 4D
architectures. A comparative alignment study characterises three strategies for lifting
per-frame reconstructions into a coherent 4D representation, establishing pairwise
hierarchical alignment as the best balance of stability and cost. Confidence-Weighted
Temporal Smoothing eliminates high-frequency jitter and reduces mean jitter by
60--91\,\% across four backbones while preserving geometric fidelity, a trade-off
native 4D architectures consistently fail to achieve.

The broader finding is that the decoupled paradigm is a principled design choice, not a workaround (see Section~\ref{sec:decoupled_paradigm}).
Geometric accuracy and temporal consistency conflict only when temporal constraints
penalise motion that is real; enforcing the separation explicitly, rather than
learning it implicitly as native 4D architectures attempt, makes the approach
robust to the sparse-view conditions where joint architectures degrade. The
results show that a strong per-frame spatial estimator with targeted
post-processing Pareto-dominates end-to-end joint architectures on the
geometry--jitter trade-off at every view count evaluated.

The remaining open problem is drift: the slow accumulation of pose estimation errors
that the low-pass filter cannot correct by design. Addressing drift through globally
coupled anchor-constrained optimisation, and extending the pipeline to close the
2-view to 4-view performance gap through reprojection-error refinement, are the natural
next steps. Both directions build directly on the framework established here and require
no additional training data, a property that made the current contribution practical
and that should guide the continuation of this work.